Concurrent Programming with CXL

### 2020-06-16 book/paper.tex
Started working on a glossary

\newglossaryentry{macro step}
{
  name=macro step,
  description={is a programmable machine that receives input,
               stores and manipulates data, and provides
               output in a useful format}
}

@@

Programming with concurrency is hard.  On the one hand concurrency
can make programs faster than sequential ones, but having multiple
processes concurrently read and update shared variables
\index{shared variable}
and synchronize with one another makes programs more complicated than
programs where only one thing happens at a time.  Why are concurrent
programs more complicated than sequential ones?
There are, at least, two reasons:
\begin{itemize}
\item The execution of a sequential
\index{sequential}
program is mostly \emph{deterministic}.
\index{determinism}
If you run it twice with the same input, the same output will be produced.
Bugs are typically easily reproducible and easy to track down, for example
by instrumenting the program.
The output of running concurrent programs depends on how the various
processes are interleaved.  Some bugs may occur only occasionally and
may never occur when the program is instrumented to find them
(so-called \emph{Heisenbugs}).
\index{Heisenbug}
\item In a sequential program, each statement and each function can be
thought of as happening \emph{atomically}
\index{atomicity}
because there is no other activity
interfering with their execution.  Even though a statement or function may
be compiled into multiple machine instructions, they are executed back-to-back
until completion.  Not so with a concurrent program, where other instructions
may update memory locations as a statement or function is being executed.
\end{itemize}
The lack of determinism and atomicity in concurrent programs make them
not only hard to reason about, but also hard to test.
Running the same test of concurrent code twice is likely to produce
two different results.  More problematically, a test may trigger a
bug only for certain ``lucky'' executions.  Due to the probabilistic
nature of concurrent code, some bugs may be highly unlikely to get
triggered even when running a test millions of times.  And even if
a bug does get triggered, the source of the bug may be hard to find
because it is hard to reproduce.

This book is intended to help people with understanding and
developing concurrent code.  In particular, it uses a new tool
called CXL that helps with \emph{testing} concurrent code.
The approach is based on \emph{model checking}:
\index{model checking}
instead of relying
on luck, CXL will run \emph{all possible executions} of a particular
test program.  So even if a bug is unlikely to occur, if the test
\emph{can} expose the bug it \emph{will}.  Importantly, if the bug is
found, the model checker precisely shows how to trigger it in
the smallest number of steps.

@@

While model checking does not in general prove an algorithm correct,
it can help with proving an algorithm correct.
The reason is that the correctness of many programs is built around
\emph{invariants}:
\index{invariant}
predicates that must hold for every state in the
execution of a program.  A model checker can find violations of
proposed invariants when evaluating a model and provide valuable early
feedback to somebody who is trying to construct a proof, even an
informal one.
Throughout this book we will include examples
of such invariants as they often provide excellent insight into
why a particular algorithm works.

@@

Concurrent programming, aka multithreaded programming, involves multiple
processes
\index{process}
running in parallel while sharing variables.
Figure~\ref{fig:inc} presents a simple example.
The program
initializes two shared variables: an integer \texttt{count} and
an array \texttt{done} with two booleans.
Method \texttt{incrementer} takes a parameter called \texttt{self}.
It increments \texttt{count} and sets \texttt{done[self]} to \texttt{True}.
Method \texttt{main} waits for the processes to finish by polling
both \texttt{done} flags.
After that, it verifies that the value of \texttt{count} equals 2.  If not,
it reports the actual value of \texttt{count}.
The program spawns three processes.
\index{process}
The first runs \texttt{incrementer(0)}, the second runs
\texttt{incrementer(1)}, and the last runs \texttt{main()}.

@@

What is going on is that the CXL program is compiled to machine instructions,
\index{machine instruction}
and it is the machine instructions that are executed by the underlying CXL
machine.  The details of this appear in Chapter~\ref{ch:cxlmachine},
but suffice it to
say that the machine has instructions that load values from memory and store
values into memory.  Importantly, it does not have instructions to atomically
increment or decrement values in memory locations.
So to increment a value in memory,
the machine must do at least three machine instructions.  Conceptually:
\begin{enumerate}
\item load the value from the memory location;
\item add 1 to the value;
\item store the value to the memory location.
\end{enumerate}

@@

Why does it work?  We will focus here on how one might go about proving
mutual exclusion for an algorithm such as Peterson's.
For that, we have to understand a little bit more about how the CXL
machine works.
In Chapter~\ref{ch:cxlmachine} we talked about the concept of \emph{state}:
\index{state}
at any point of time the CXL machine is in a specific state.
A state is comprised of the values of the shared variables as well as
the values of the process variables for each process, including its
program counter and the contents of its stack.
Everytime a process executes a CXL machine instruction, the
state changes (if only because the program counter of the process
changes).  We call that a \emph{step}.
\index{step}
Steps in CXL are atomic.

@@

A \emph{micro step}
\index{micro step}
is the execution of a single CXL machine instruction
by a context.
Each micro step generates a new state.
When there are multiple contexts, the CXL machine can interleave them.
However, trying to interleave every microstep would be needlessly expensive,
as many micro steps involve changes to a context that are invisible to
other contexts.

@@

% \glsaddall
\printglossaries

>>>>

\newglossaryentry{atomicity}
{
  name=atomicity,
  description={describes that a certain machine instruction or sequence of machine
    instructions by a process is executed indivisibly and cannot be interleaved with 
    machine instructions of another process}
}
\newglossaryentry{concurrent execution}
{
  name=concurrent execution,
  description={is when there are multiple processes executing and their machine
    instructions are interleaved in an unpredictable manner}
}
\newglossaryentry{determinism}
{
  name=determinism,
  description={is when the outcome of an execution is uniquely determined by the
    initial state}
}
\newglossaryentry{invariant}
{
  name=invariant,
  description={is a binary property that must hold for every state of a process}
}
\newglossaryentry{machine instruction}
{
  name=machine instruction,
  description={is an atomic operation on the CXL machine, executed by a process}
}
\newglossaryentry{macro step}
{
  name=macro step,
  description={is a sequence of micro steps with at most one instruction
        externally visible to other processes.  It is the unit of interleaving
        in CXL}
}
\newglossaryentry{micro step}
{
  name=micro step,
  description={is the execution of one machine instruction by a process, updating
        its state and the state of the CXL machine}
}
\newglossaryentry{model checking}
{
  name=model checking,
  description={is a formal verification method that explores all possible executions
    of a program, which must have a finite number of states}
}
\newglossaryentry{process}
{
  name=process,
  description={is a method in execution.  We do not make the distinction
      between processes and threads.  A process has a program counter, a register,
      and a stack}
}
\newglossaryentry{process variable}
{
  name=process variable,
  description={is a variable that is private to a single process and stored in
        its register}
}
\newglossaryentry{sequential execution}
{
  name=sequential execution,
  description={is when there is just one process executing, as opposed to concurrent
    execution}
}
\newglossaryentry{shared variable}
{
  name=shared variable,
  description={is a variable that is stored in the memory of the CXL machine and
      shared between multiple processes, as opposed to a process variable}
}

@@

Programming with concurrency is hard.  On the one hand concurrency
can make programs faster than sequential ones, but having multiple
processes (aka \emph{threads}\footnote{We
\index{thread}
do not make a distinction between processes and threads.}) concurrently
\glsadd{concurrent execution}
read and update shared variables
\index{shared variable}
\glsadd{shared variable}
and synchronize with one another makes programs more complicated than
programs where only one thing happens at a time.  Why are concurrent
programs more complicated than sequential ones?
There are, at least, two reasons:
\begin{itemize}
\item The execution of a sequential
\index{sequential}
\glsadd{sequential execution}
program is mostly \emph{deterministic}.
\index{determinism}
\glsadd{determinism}
If you run it twice with the same input, the same output will be produced.
Bugs are typically easily reproducible and easy to track down, for example
by instrumenting the program.
The output of running concurrent programs depends on how the various
processes are interleaved.  Some bugs may occur only occasionally and
may never occur when the program is instrumented to find them
(so-called \emph{Heisenbugs}).
\index{Heisenbug}
\item In a sequential program, each statement and each function can be
thought of as happening \emph{atomically}
\index{atomicity}
\glsadd{atomicity}
because there is no other activity
interfering with their execution.  Even though a statement or function may
be compiled into multiple machine instructions, they are executed back-to-back
until completion.  Not so with a concurrent program, where other instructions
may update memory locations as a statement or function is being executed.
\end{itemize}
The lack of determinism and atomicity in concurrent programs make them
not only hard to reason about, but also hard to test.
Running the same test of concurrent code twice is likely to produce
two different results.  More problematically, a test may trigger a
bug only for certain ``lucky'' executions.  Due to the probabilistic
nature of concurrent code, some bugs may be highly unlikely to get
triggered even when running a test millions of times.  And even if
a bug does get triggered, the source of the bug may be hard to find
because it is hard to reproduce.

This book is intended to help people with understanding and
developing concurrent code.  In particular, it uses a new tool
called CXL that helps with \emph{testing} concurrent code.
The approach is based on \emph{model checking}:
\index{model checking}
\glsadd{model checking}
instead of relying
on luck, CXL will run \emph{all possible executions} of a particular
test program.  So even if a bug is unlikely to occur, if the test
\emph{can} expose the bug it \emph{will}.  Importantly, if the bug is
found, the model checker precisely shows how to trigger it in
the smallest number of steps.

@@

While model checking does not in general prove an algorithm correct,
it can help with proving an algorithm correct.
The reason is that the correctness of many programs is built around
\emph{invariants}:
\index{invariant}
\glsadd{invariant}
predicates that must hold for every state in the
execution of a program.  A model checker can find violations of
proposed invariants when evaluating a model and provide valuable early
feedback to somebody who is trying to construct a proof, even an
informal one.
Throughout this book we will include examples
of such invariants as they often provide excellent insight into
why a particular algorithm works.

@@

Concurrent programming, aka multithreaded programming, involves multiple
processes
\index{process}
\glsadd{process}
running in parallel while sharing variables.
Figure~\ref{fig:inc} presents a simple example.
The program
initializes two shared variables: an integer \texttt{count} and
an array \texttt{done} with two booleans.
Method \texttt{incrementer} takes a parameter called \texttt{self}.
It increments \texttt{count} and sets \texttt{done[self]} to \texttt{True}.
Method \texttt{main} waits for the processes to finish by polling
both \texttt{done} flags.
After that, it verifies that the value of \texttt{count} equals 2.  If not,
it reports the actual value of \texttt{count}.
The program spawns three processes.
The first runs \texttt{incrementer(0)}, the second runs
\texttt{incrementer(1)}, and the last runs \texttt{main()}.

@@

What is going on is that the CXL program is compiled to machine instructions,
\index{machine instruction}
\glsadd{machine instruction}
and it is the machine instructions that are executed by the underlying CXL
machine.  The details of this appear in Chapter~\ref{ch:cxlmachine},
but suffice it to
say that the machine has instructions that load values from memory and store
values into memory.  Importantly, it does not have instructions to atomically
increment or decrement values in memory locations.
So to increment a value in memory,
the machine must do at least three machine instructions.  Conceptually:
\begin{enumerate}
\item load the value from the memory location;
\item add 1 to the value;
\item store the value to the memory location.
\end{enumerate}

@@

Why does it work?  We will focus here on how one might go about proving
mutual exclusion for an algorithm such as Peterson's.
For that, we have to understand a little bit more about how the CXL
machine works.
In Chapter~\ref{ch:cxlmachine} we talked about the concept of \emph{state}:
\index{state}
at any point of time the CXL machine is in a specific state.
A state is comprised of the values of the shared variables as well as
the values of the process variables
\index{process variable}
\glsadd{process variable}
for each process, including its
program counter and the contents of its stack.
Everytime a process executes a CXL machine instruction, the
state changes (if only because the program counter of the process
changes).  We call that a \emph{step}.
\index{step}
Steps in CXL are atomic.

@@

A \emph{micro step}
\index{micro step}
\glsadd{micro step}
is the execution of a single CXL machine instruction
by a context.
Each micro step generates a new state.
When there are multiple contexts, the CXL machine can interleave them.
However, trying to interleave every microstep would be needlessly expensive,
as many micro steps involve changes to a context that are invisible to
other contexts.

@@

\printglossaries
